WORCS: A Workflow for Open Reproducible Code in Science

### 2020-11-05 paper/Manuscript_DS.Rmd
R1 C7a Big data

  affiliation: | 
      a: Utrecht University faculty of Social and Behavioral Sciences, department of Methodology & Statistics  
      b: Open Science Community Utrecht  
      c: Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany  
      d: Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany and London, UK  
      e: UMC Utrecht  
      f: Utrecht University, Department of Information and Computing Sciences  
      g: Humboldt-Universitaet zu Berlin  
      h: Utrecht University faculty of Humanities, department of Language, Literature & Communication, Utrecht Institute of Linguistics OTS  
      i: University Library Utrecht
title: "WORCS: A Workflow for Open Reproducible Code in Science"
short:      "WORKFLOW FOR OPEN SCIENCE"
shorttitle:      "WORKFLOW FOR OPEN SCIENCE"
abstract: |
  Adopting open science
  principles can be challenging and time-intensive, because doing so requires
  substantial conceptual education and training in the use of new tools.
  This paper introduces the Workflow for
  Open Reproducible Code in Science (WORCS): A step-by-step
  procedure that researchers can follow to make a research project open and
  reproducible.
  The purpose of the workflow is to lower the threshold for 
  adoption of open science principles.
  It is based on best practices, and can be used either in
  parallel to, or in absence of, top-down requirements by journals,
  institutions, and funding bodies.
  To facilitate widespread adoption, the WORCS principles have been implemented
  in the R package `worcs`, which offers an RStudio 
  project template and utility functions for specific workflow steps.
  This paper introduces the conceptual workflow,
  discusses how it meets different 
  standards for open science, and provides examples of the implementation in `R`. 
  This paper is primarily targeted towards
  scholars conducting research projects in R on a private computer,
  conducting research that involves academic prose, analysis code,
  and tabular data that does not require specialized computational
  provisions.
  However, the workflow is flexible enough to accommodate other scenarios,
  and offers a sensible starting point for customized solutions.
  The source code for the `R` package and manuscript,
  and a list of user examples of WORCS projects, are available at 
  https://github.com/cjvanlissa/worcs.
authornote: "blabla"
keywords:
  # at least one keyword must be supplied
  formatted: [open science, reproducibility, r, dynamic document generation, version control]
  plain:     [open science, reproducibility, r, dynamic document generation, version control]
preamble: >
  \usepackage{amsmath}
bibliography      : ["worcs.bib"]  
urlcolor          : blue
output: rmarkdown::html_document
---
<!--
, "worcs_packages.bib"
output: 
  rmarkdown::html_document:
    citation_package: biblatex Note: Citations are always handled by pandoc when format is not PDF
knit: worcs::cite_all
csl: apa.csl
csl: apa.csl
-->

@@

Data sharing is important for computational reproducibility and secondary
analysis. Computational reproducibility means that a third party can exactly
recreate the results from the original data, using the published analysis code.
Secondary analysis means that a third party can conduct sensitivity analyses,
explore alternative explanations, or even use existing data to answer a
different research question. From an open science perspective, data sharing is
always desirable. From a practical point of view, it is not always possible.
Data sharing may be impeded by lack of storage space, legal constraints, ethical restrictions, absence
of informed consent for data sharing, or privacy concerns. For example, one
concern that applies when working with human participant data, is
pseudonimization. 
Many legislatures require researchers to 'de-identify' human subject data upon collection, by removing or deleting any sensitive personal information and
contact details. However, when researchers have an imperfect understanding of the
legal obligations and dispensations, they can end up over- or undersharing, and
risk either placing participants at risk of being identified, or undermine the
replicability of their own work and potential reuse value of their data [@phillipsDiscombobulationDeidentification2016].

Once data are de-identified or pseudononimized, they can often be shared if participants have
provided informed consent to that effect. Nevertheles, it is important to note
that the European GDPR prohibits storing 'personal data' (information which can
identify a natural person whether directly or indirectly) on a server outside
the EU, unless it offers an "adequate level of protection". Although different
rules may apply to pseudonimized data, there are many repositories that are
GDPR compliant, such as the European servers of the Open Science Framework.
Different Universities, countries, and funding bodies also have their own
repositories that are complient with local legistation. Before sharing any
human participant data, it is recommended to obtain approval from an internal
(ethical) review board, guidance from Research Data Management Support, and
informed consent from participants.

If data can be shared openly, they can simply be pushed to a public Git
repository, along with the analysis code. This way, others can download the
entire repository and reproduce the analyses from start to finish. 
If data cannot be shared, researchers should aim to safeguard the potential for
computational reproducibility and secondary analysis as much as possible. WORCS
recommends two solutions to accomplish this goal.
The first solution is to publish a *checksum* of the original data file [@rivest1992md5].
Think of a checksum as a 32-character summary, or as a "one word" description,
of the contents of a file.
Any change to the file will result in a different checksum^[It is theoretically possible but improbable that random changes to a file will result in the same checksum.]. 
Thus, one can use a checksum to verify the identity of a file, and ensure that its contents are unchanged.
When data cannot be shared, the risk of fraud or abuse can be mitigated by publishing a checksum for the original data, in addition to the complete analysis code.
Using the checksum to verify the identity of a private dataset, researchers can prove to an independent auditor that running the public analysis code on the private data results in the published results of their work.

The second solution is to share a synthetic dataset with similar characteristics
to the real data. Synthetic data mimic the level of measurement and
(conditional) distributions of the real data
[see @nowokSynthpopBespokeCreation2016]. Sharing synthetic data allows any third
party to 1) verify that the published code works, 2) debug the code, and 3)
write valid code for alternative analyses. It is important to note that complex
multivariate relationships present in the real data are often lost in synthetic
data. Thus, findings from the real data might not be replicated in the synthetic
data, and findings in the synthetic data *should not be substantively interpreted*.
Still, sharing synthetic data facilitates data requests from third parties. A
third party can write analysis code based on the synthetic data, and send it to
the authors who evaluate it on the real data and send back the results. The
`worcs` package offers a simple but flexible function to generate synthetic data,
`synthetic()`, which by default uses random forests to generate a model for
synthesis (see @wrightRangerFastImplementation2015).
This function is based on the work by @nowokSynthpopBespokeCreation2016 (see the `R` package `synthpop`),
but with a simpler and more flexible user interface.
Specifically, the `worcs` function `synthetic()` accepts any function that can be used to model
marginal associations in the data.
For further details, see the [function documentation](https://cjvanlissa.github.io/worcs/reference/synthetic.html).

@@

>>>>

  affiliation: | 
      a: Utrecht University faculty of Social and Behavioral Sciences, department of Methodology & Statistics  
      b: Open Science Community Utrecht  
      c: Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany  
      d: Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany and London, UK  
      e: UMC Utrecht  
      f: Utrecht University, Department of Information and Computing Sciences  
      g: Humboldt-Universitaet zu Berlin  
      h: Utrecht University faculty of Humanities, department of Language, Literature & Communication, Utrecht Institute of Linguistics OTS  
      i: University Library Utrecht
title: "WORCS: A Workflow for Open Reproducible Code in Science"
short:      "WORKFLOW FOR OPEN SCIENCE"
shorttitle:      "WORKFLOW FOR OPEN SCIENCE"
abstract: |
  Adopting open science
  principles can be challenging and time-intensive, because doing so requires
  substantial conceptual education and training in the use of new tools.
  This paper introduces the Workflow for
  Open Reproducible Code in Science (WORCS): A step-by-step
  procedure that researchers can follow to make a research project open and
  reproducible.
  The purpose of the workflow is to lower the threshold for 
  adoption of open science principles.
  It is based on best practices, and can be used either in
  parallel to, or in absence of, top-down requirements by journals,
  institutions, and funding bodies.
  To facilitate widespread adoption, the WORCS principles have been implemented
  in the R package `worcs`, which offers an RStudio 
  project template and utility functions for specific workflow steps.
  This paper introduces the conceptual workflow,
  discusses how it meets different 
  standards for open science, and provides examples of the implementation in `R`. 
  This paper is primarily targeted towards
  scholars conducting research projects in R on a private computer,
  conducting research that involves academic prose, analysis code,
  and (optionally) tabular data.
  However, the workflow is flexible enough to accommodate other scenarios,
  and offers a sensible starting point for customized solutions.
  The source code for the `R` package and manuscript,
  and a list of user examples of WORCS projects, are available at 
  https://github.com/cjvanlissa/worcs.
authornote: "blabla"
keywords:
  # at least one keyword must be supplied
  formatted: [open science, reproducibility, r, dynamic document generation, version control]
  plain:     [open science, reproducibility, r, dynamic document generation, version control]
preamble: >
  \usepackage{amsmath}
bibliography      : ["worcs.bib"]  
urlcolor          : blue
output: rmarkdown::html_document
---
<!--
, "worcs_packages.bib"
output: 
  rmarkdown::html_document:
    citation_package: biblatex Note: Citations are always handled by pandoc when format is not PDF
knit: worcs::cite_all
csl: apa.csl
csl: apa.csl
-->

@@

Data sharing is important for computational reproducibility and secondary
analysis. Computational reproducibility means that a third party can exactly
recreate the results from the original data, using the published analysis code.
Secondary analysis means that a third party can conduct sensitivity analyses,
explore alternative explanations, or even use existing data to answer a
different research question. 
If it is possible to share the data, these can simply be pushed to a Git remote
repository, along with documentation (such as a codebook) and the analysis code.
This way, others can download the
entire repository and reproduce the analyses from start to finish. 
From an open science perspective, data sharing is
always desirable. From a practical point of view, it is not always possible.

Data sharing may be impeded by several concerns.
One technical concern is whether data require special storage provisions,
as is often the case with "big data".
Most Git remote repositories do not support files larger than 100MB,
but it is possible to enhance a Git repository with [Git Large File Storage](https://git-lfs.github.com/) to add remotely stored files as large as several GB.
Files exceeding this size require custom solutions beyond the scope of this paper.

Sharing human participant data additionally entails legal and ethical concerns.
Before sharing any
human participant data, it is recommended to obtain approval from qualified
ethical and legal advisory organs, guidance from Research Data Management Support, and
informed consent from participants.
Many legislatures require researchers to 'de-identify' human subject data upon collection, by removing or deleting any sensitive personal information and
contact details. However, when researchers have an imperfect understanding of the
legal obligations and dispensations, they can end up over- or undersharing, and
risk either placing participants at risk of being identified, or undermine the
replicability of their own work and potential reuse value of their data [@phillipsDiscombobulationDeidentification2016].
Furthermore, it is important to note
that the European GDPR prohibits storing 'personal data' (information which can
identify a natural person whether directly or indirectly) on a server outside
the EU, unless it offers an "adequate level of protection". Although different
rules may apply to pseudonimized data, there are many repositories that are
GDPR compliant, such as the European servers of the Open Science Framework.
Different Universities, countries, and funding bodies also have their own
repositories that are complient with local legistation.
In sum, data should only be shared after consult qualified legal and ethical advisory organs,
obtaining informed consent from participants,
and de-identifying or pseudononimizing data.

If data cannot be shared, researchers should aim to safeguard the potential for
computational reproducibility and secondary analysis as much as possible. WORCS
recommends two solutions to accomplish this goal.
The first solution is to publish a *checksum* of the original data file [@rivest1992md5].
Think of a checksum as a 32-character unique identifier^[It is theoretically possible but improbable that random changes to a file will result in the same checksum.],
or as a lossy "summary" of the contents of a file.
Any change to the file will result in a different checksum. 
Thus, one can use a checksum to verify the identity of a file, and ensure that its contents are unchanged.
When data cannot be shared, the risk of fraud or abuse can be mitigated by publishing a checksum for the original data, in addition to the complete analysis code.
Using the checksum to verify the identity of a private dataset, researchers can prove to an independent auditor that running the public analysis code on the private data results in the published results of their work.

A second solution is to share a synthetic dataset with similar characteristics
to the real data. Synthetic data mimic the level of measurement and
(conditional) distributions of the real data
[see @nowokSynthpopBespokeCreation2016]. Sharing synthetic data allows any third
party to 1) verify that the published code works, 2) debug the code, and 3)
write valid code for alternative analyses. It is important to note that complex
multivariate relationships present in the real data are often lost in synthetic
data. Thus, findings from the real data might not be replicated in the synthetic
data, and findings in the synthetic data *should not be substantively interpreted*.
Still, sharing synthetic data facilitates data requests from third parties. A
third party can write analysis code based on the synthetic data, and send it to
the authors who evaluate it on the real data and send back the results. The
`worcs` package offers a simple but flexible function to generate synthetic data,
`synthetic()`, which by default uses random forests to generate a model for
synthesis (see @wrightRangerFastImplementation2015).
This function is based on the work by @nowokSynthpopBespokeCreation2016 (see the `R` package `synthpop`),
but with a simpler and more flexible user interface.
Specifically, the `worcs` function `synthetic()` accepts any function that can be used to model
marginal associations in the data.
For further details, see the [function documentation](https://cjvanlissa.github.io/worcs/reference/synthetic.html).

@@

Another limitation is the fact that no single workflow can suit all research projects.
Not all projects conform to the steps outlined here;
some may skip steps, add steps, or follow similar steps in a different order.
In principle, WORCS does not enforce a linear order, and allows extentensive user flexibility.
A related concern is that some projects might require specialized solutions outside of the scope of this paper.
For example, some projects might use very large data files,
rely on data that reside on a protected server,
or use proprietary software dependencies that cannot be version controlled using `renv` or Docker.
In such cases, the workflow can serve as a starting point for a custom approach;
offering a primer on best practices, and a discussion of relevant factors to consider.
We urge users to inform us of such challenges and custom solutions,
so that we may add them to the [overview of WORCS-projects on GitHub](https://github.com/cjvanlissa/worcs), where they can serve as an example to others.

### 2020-11-05 paper/reviews_ds.Rmd
R1 C7a Big data

We agree that GitHub is not an ideal platform for handling data larger than 100MB.
However, 100MB is very large for tabular data; for example, if only integer numbers are stored that is $(100MB*1024^2)/(2^2) = 26214400$ unique values.

GitHub is, however, ideal version controlling analysis code, RMarkdown documents, and small to medium-sized data files.
Larger files can be linked to git using [git-annex](https://git-annex.branchable.com),
[Git Large File Storage](https://git-lfs.github.com/), and similar solutions. 
We now mention git-annex as a possible route to handle larger data files (TODO).

>>>>

Response: We have split this comment up to separately address the handling of large data, and findability (see below).

With regard to the handling of large data,
we agree that GitHub is not an ideal platform for handling data larger than 100MB.
It is important to note, however, that there are no restrictions for sharing files up to that size.
Further, take into consideration that 100MB is very large for tabular data;
for example, if only integer numbers are stored that is $(100MB*1024^2)/(2^2) = 26214400$ unique values.
In the social sciences, datasets rarely exceed this limit -
and so for our readership,
GitHub is as good a platform for sharing data as any.
Moreover, GitHub is ideal for version controlling analysis code, RMarkdown documents, and small to medium-sized data files.
In our opinion, the advantages of having **all** of a paper's resources in the same repository outweighs the advantages of encouraging every reader to additionally use specialized solutions for "big data" (which most will not need).

To address this comment, we now explicitly address the problem of version controlling large files, and introduce a solution for larger files, which can be linked to git using
[Git Large File Storage](https://git-lfs.github.com/):

> Data sharing may be impeded by several concerns.
One technical concern is whether data require special storage provisions,
as is often the case with "big data".
Most Git remote repositories do not support files larger than 100MB,
but it is possible to enhance a Git repository with [Git Large File Storage](https://git-lfs.github.com/) to add remotely stored files as large as several GB.
Files exceeding this size require custom solutions beyond the scope of this paper.

We also added a paragraph to the Limitations section that discusses the fact that no single workflow is suitable for all situations:

> Another limitation is the fact that no single workflow can suit all research projects.
Not all projects conform to the steps outlined here;
some may skip steps, add steps, or follow similar steps in a different order.
In principle, WORCS does not enforce a linear order, and allows extentensive user flexibility.
A related concern is that some projects might require specialized solutions outside of the scope of this paper.
For example, some projects might use very large data files,
rely on data that reside on a protected server,
or use proprietary software dependencies that cannot be version controlled using `renv` or Docker.
In such cases, the workflow can serve as a starting point for a custom approach;
offering a primer on best practices, and a discussion of relevant factors to consider.
We urge users to inform us of such challenges and custom solutions,
so that we may add them to the [overview of WORCS-projects on GitHub](https://github.com/cjvanlissa/worcs), where they can serve as an example to others.

\Assignment{BV, AB, AL, CJ}
\WorkInProgress
\Done
